---
title: "Access to gaia server"
author: "Roland Krause"
date: "15 January 2018"
output: 
  iosp::ioslides_plus
---

```{r include=FALSE}
library(tidyverse)
```
# Setting 

## Content 

### Basic set-up {.bg-green .box-8 .offset-2}
 * SSH keys
 * Virtual machines
 * Server 
 * [HPC](hpc.uni.lu)
 

## Access to Gaia server {.smaller}
### Access via PuTTY {.bg-blue .box-12}
 * Windows machines
 * Convert given key to putty format (stored in `Documents` folder) using PuTTYgen
 * Open PuTTY
 * Give the key under SSH - Auth
 * Connect to student`<no>`@access-gaia.uni.lu with Port 8022



### SSH access via Shell {.bg-yellow .box-12}
 * From linux/mac 
 * Virtual machine under Windows

```{bash eval=FALSE}
ssh –p 8022 –i $HOME/.ssh/student<no>.key student<no>@access-gaia.uni.lu

```




## Access node, scheduling and compute node 

### Basic cluster architecture {.bg-yellow .offset-2 .box-10}
 * Access node
    - For job submission only. 
    - Do not use for any computation
    - Do not use for CPU intensive jobs, e.g. unzip, compilation
    - Only basic framework installed
 * Compute nodes - Run jobs
    - Enclosure, node, cpu, core
 * Scheduler
    - Never access by normal users, distributes task to compute clusters 
 
## The OAR scheduler

### Submission {.bg-blue .box-10 .offset-1}

[OAR explained](https://hpc.uni.lu/users/docs/oar.html)

On `access-gaia`
```{bash eval=FALSE, title="OAR submission"}

oarsub "my command"
oarsub -I # Interactive session
oarsub myscript.bash

```

## Reserved container

```{bash eval=F, title="For 2018-01 only"}
oarsub -t inner=<container job id> –lcore=4,walltime=12 
oarsub -t inner=4274331 -lcore=1,walltime=12  # 2018-01-15 through 2018-01-20 only


```


## Copy data from gaia to your local computer | Mac/Linux systems
```{bash eval=FALSE}

scp –P 8022 –r –i $HOME/.ssh/student<no>.key \
student<no>@access-gaia.uni.lu:<path-to-file/dir><destination>

scp –P 8022 –r –i student<no>.key student<no>@access-gaia.uni.lu:~/WEScourse.tar.gz .

export LC_ALL=en_US.UTF-8

```



## Copy data from gaia to your local computer | Windows

### WinSCP {.bg-green .offset-1 .box-10}
* Install WinSCP (in case it's not set-up yet)
* Open WinSCP 
    + File protocol `SCP`
    + Hostname `access-gaia.uni.lu`
    + Username student<no> i.e. `student13`
    + Advanced: SSH - Authentification. 
      Give Key student<no>.ppk, typically saved to Documents folder
* Use PuTTY instructions
* Open terminal windows


## Worked example

### Reading an Variant Call Format file produced on the server into R  {.bg-green .box-12}

Create a directory to work in your home directory.

```{bash eval=F}

mkdir ~/work_isb/

scp -P 8022 -i ~/.ssh/student<no>.key \
student13@access-gaia.uni.lu:~/unixtut/media/Mills_and_1000G_gold_standard.indels.b37.chr22.vcf \ ~/work_isb/ 

```


## Variant Call Format 
```{r title="Reading in a VCF file"}

my_vcf <- read_delim("~/work_isb/Mills_and_1000G_gold_standard.indels.b37.chr22.vcf", delim="\t", comment="#", col_names = c("CHROM",  "POS","ID" ,"REF"  , "ALT", "QUAL" , "FILTER",   "INFO"))

```